Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2018-11-05 doc/network.tex
update

An epidemic can be characterized by its speed (exponential growth rate, $r$) and its strength (reproductive number, \RR).
Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic [CITE].
However, directly measuring the reproductive number often requires detailed knowledge of disease natural history and may not be feasible early in an epidemic [CITE].
Instead, the reproductive number can be \emph{indirectly} estimated from exponential growth rate, which can easily be estimated from incidence data [CITE].
These two quantities are linked by generation-interval distributions \citep{wallinga2007generation}.

@@

\section{Inferring generation-interval distributions from contact tracing data}

When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric effect, (3) local spatial effect and (4) global spatial effect.
Right-censoring effect and egocentric effect are well-understood and can be corrected for.
While local and global spatial effects are difficult to estimate, we can make qualitative predictions about their effects on the realized generation intervals and reproductive number. 
Both local and global effects reduce number of infections that occur and shorten generation intervals but the time scale on which they matter differ:
local spatial effect will be important early in an epidemic whereas the global effect will be small until after the exponential growth period has ended.

It is clear that aggregating generation intervals through space and time yields a distribution that is different from the intrinsic distribution.
When the epidemic is growing, the proportion of susceptible individuals remain approximately constant and the intrinsic (or conditional) generation-interval distributions can be inferred from the observed intervals.
In this section, we investigate two statistical methods for recovering the intrinsic (or conditional) generation-interval distribution.
Detailed derivations of the two methods can be found in section.

\swp{maybe call it something else besides nonparametric method; it's still possible to make assumptions about the distribution and make inference}
We refer to the first method as the population-level method as it relies on the observed distribution without taking into account who infected whom.
As the observed generation-interval distribution is a weighted distribution of the intrinsic distribution, the intrinsic distribution can be recovered by taking the inverse of the weights, given by \eref{exp}.
Hence, the population-level method relies on the observed generation intervals and the exponential growth rate:
\begin{equation}
g(\tau) \propto \tsub{c}{exp}(\tau) \exp(r\tau).
\end{equation}

While the non-parametric method is simple, it does not use all available information from contact tracing data: it does not take into account who infected whom.
Assuming that infection occur as a poisson process in the beginning of an epidemic, we can obtain the following likelihood for the observed infection process:
\begin{equation}
\mathcal{L}(\RR, \theta) = \prod_{j=1}^N \left(\RR^{n_j} \exp \left(- \RR \int_0^{\tsub{t}{censor} - t_j} g(s;\theta) ds \right) \prod_{i=1}^{n_j} g(\tau_{i, j}; \theta) \right),
\end{equation}
where $N$ is the total number of infected individuals, $n_j$ is the number of individuals infected by $j$, $\tsub{t}{censor}$ is the time period which censoring was performed until, $t_j$ is the time at which infector $j$ was infected, $\tau_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is the (vector) parameter of the generation interval distribution $g$.
A similar method was proposed by \cite{forsberg2008likelihood} but it relies on discretized incidence reports rather than the observed generation intervals.

@@

First, we compare how the estimates from the two methods vary across time using a single stochastic simulation on a homogeneous network (Figure~\ref{fig:example}).
During the initial exponential growth period (approximately between 9 and 20 weeks), both nonparametric and parametric methods yield similar estimates of mean generation interval and reproductive number.
However, we observe opposite trends after the growth period (approximately after 20 weeks).
As the nonparametric method takes a weighted average of the observed distribution, increase in the observed mean generation interval leads to increase in the nonparametric estimates of both mean generation interval and reproductive number.
On the other hand, as the parametric method assumes a Poisson process with constant proportion of susceptibles, decrease in the number of susceptible results in underestimation of both mean generation interval and reproductive number.
we observe grdaual increase in the observed mean-generation interval as we aggregate more data, but the observed estimates always underestimate both mean generation interval and reproductive number, as predicted from the deterministic simulation.

@@

Then, we compare accuracy and reliability of the two methods during the initial growth period using 100 stochastic simulations on a homogeneous network (Figure~\ref{fig:test}).
We find that the parametric method provides the most consistent (least variable) estimates.
The nonparametric method overestimates reproductive number initially but this is likely driven by the initial overestimation of the growth rate (Figure~\ref{fig:example}).

\swp{May need to explain what coverage is...}
Surprisingly, both methods fail to attain nominal coverage;
throughout the estimation period, both methods attain approximately 70\% coverage.
Even though the gamma distribution looks indistinguishable from the shape of the intrinsic generation-interval distribution, making a wrong distributional assumption leads to underconfidence.
To confirm that the undercoverage is caused by distribuional assumption, we simulated an epidemic using gamma kernel and found that using the true shape yields nominal coverage (see Appendix).
These results are particularly alarming because shape of the generation-interval distribution have often been assumed in outbreak analyses and it is impossible to know the true shape of the distribution [CITE].
\swp{Why does nonparametric method undercover?}

>>>>

An epidemic can be characterized by its speed (exponential growth rate, $r$) and its strength (reproductive number, \RR).
Reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic [CITE].
However, directly measuring the reproductive number often requires detailed knowledge of disease natural history and may not be feasible early in an epidemic [CITE].
Instead, the reproductive number can be \emph{indirectly} estimated from exponential growth rate, which can easily be estimated from incidence data \citep{mills2004transmissibility, nishiura2009transmission, ma2014estimating}.
These two quantities are linked by generation-interval distributions \citep{wallinga2007generation}.

@@

\section{Inferring generation-interval distributions from a contact tracing data}

When generation intervals are sampled through contact tracing, there will be four effects present in the sample: (1) right-censoring effect, (2) egocentric effect, (3) local spatial effect and (4) global spatial effect.
Right-censoring effect and egocentric effect are well-understood.
While local and global spatial effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive number. 
Both local and global effects reduce number of infections that occur and shorten generation intervals but the time scale on which they matter differs:
local spatial effect will be important early in an epidemic whereas the global effect will be small until sufficient amount of susceptible individuals become infected.
As a result, both local and egocentric effects affect the initial spread of a disease and therefore the realized reproductive number.

Since the right-censoring effect is a sampling artifact, it must be corrected for.
On the other hand, local and egocentric effects will be embedded within realized generation intervals and do not need to be corrected for as they reflect the characteristic of a realized epidemic.
We claim that accounting for the right-censoring effect is sufficient and any necessary spatial component will be accounted for implicitly.
Hereafter, we will refer to temporally corrected distribution as the effective generation-interval distribution to contrast from the intrinsic generation-interval distribution; only in a large homogeneously mixing population, the corrected generation-interval distribution will be equivalent to the intrinsic generation-interval distribution.
In this section, we investigate two statistical methods for correcting for temporal bias in a contact tracing data.
Detailed derivations of the two methods can be found in section ???.

We refer to the first method as the population-level method as it relies on the  observed distribution aggregated across population.
As the observed generation-interval distribution is a weighted distribution of the effective distribution, the effective distribution can be recovered by taking the inverse of the weights, given by \eref{exp}.
Hence, the population-level method relies on the observed generation intervals and the exponential growth rate:
\begin{equation}
g(\tau) \propto \tsub{c}{exp}(\tau) \exp(r\tau).
\end{equation}

While the population-level method is simple and intuitive, it does not use all available information from contact tracing data as it does not take into account who infected whom.
Instead, we can model the initial spread of an epidemic using a poisson process on an individual-level and obtain the following likelihod::
\begin{equation}
\mathcal{L}(\RR, \theta) = \prod_{j=1}^N \left(\RR^{n_j} \exp \left(- \RR \int_0^{\tsub{t}{censor} - t_j} g(s;\theta) ds \right) \prod_{i=1}^{n_j} g(\tau_{i, j}; \theta) \right),
\end{equation}
where $N$ is the total number of infected individuals, $n_j$ is the number of individuals infected by $j$, $\tsub{t}{censor}$ is the time period which censoring was performed until, $t_j$ is the time at which infector $j$ was infected, $\tau_{i,j}$ is the observed generation interval between infector $j$ and infectee $i$, and $\theta$ is the (vector) parameter of the generation-interval distribution $g$.
A similar method was proposed by \cite{forsberg2008likelihood} but it relies on discretized incidence reports rather than the observed generation intervals.

@@

First, we compare how the estimates from the two methods vary across time using a single stochastic simulation on a homogeneous network (Figure~\ref{fig:example}).
During the initial exponential growth period (approximately between 9 and 20 weeks), both population-level and individual-level methods yield similar estimates of mean generation interval and reproductive number.
However, we observe opposite trends after the growth period (approximately after 20 weeks).
As the population-level method takes a weighted average of the observed distribution, increase in the observed mean generation interval leads to increase in the nonparametric estimates of both mean generation interval and reproductive number.
On the other hand, as the individual-level method assumes a Poisson process with constant proportion of susceptibles, decrease in the number of susceptible is translated to underestimation of both mean generation interval and reproductive number.
we observe grdaual increase in the observed mean-generation interval as we aggregate more data, but using the raw observed generation interval distributions always underestimate both mean generation interval and reproductive number, as predicted from the deterministic simulation (Figure~\ref{fig:censor}).

@@

We compare accuracy and reliability of the two methods during the initial growth period using 100 stochastic simulations on a homogeneous network (Figure~\ref{fig:test}).
We find that the individual-level method provides the most consistent (least variable) estimates.
The population method overestimates reproductive number initially but this is likely driven by the initial overestimation of the growth rate (Figure~\ref{fig:example}).

\swp{May need to explain what coverage is...}
We also compare coverage probabilities of these methods. 
Coverage probabilities are defined as the proportion of onfidence intervals that contain the true value in a repeated simulations; for example, a 95\% confidence interval should attain 95\% coverage probability by its definition.
Surprisingly, both methods only attain approximately 70\% coverage.
Even though the gamma distribution looks indistinguishable from the shape of the intrinsic generation-interval distribution, making a wrong distributional assumption leads to narrow confidence intervals.
To confirm that the undercoverage is caused by distribuional assumption, we simulated an epidemic using gamma intrinsic generation-interval and found that fitting the true shape yields nominal coverage (see Appendix).
These results are particularly alarming because shape of the generation-interval distribution have often been assumed in outbreak analyses and it is impossible to know the true shape of the distribution [CITE].
\swp{Why does nonparametric method undercover?}
